Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-12 paper.qmd
methods
methods and materials

```{=html}
<!--

@@

-->
```

@@

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share an established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results. At the 'soft' end of the spectrum (e.g. economics and psychology), the theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. However, independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000]. Similarly, quantitative analysis of the frequency of positive results (ie. full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@PositiveResultsIncrease]. The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].

Here I take a bibliometric approach to objectively quantify the relative hardness or softness of archaeology to evaluate its status as a science, and the place of this journal in the discipline. For a large sample of peer-reviewed journal articles, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references. These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people with specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail [@zuckerman1972age]. Harder disciplines tend to have higher average numbers of authors on papers. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. I used page numbers to measure article length. Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus need citations for support, relative to harder fields [@skilton2006comparative]. The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation]. I calculated the Price index for each article which is the proportion of citated worked published in the five years preceding the paper.

n an ideal science, scholars share a common background of established theories, facts and methods. This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research. Harder sciences are hypothesised to come closer to this ideal. Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)

@@

My motivation is the view that It is customary in significant round-number anniversary

>>>>

<!--

@@

-->

@@

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results. At the 'soft' end of the spectrum (e.g. economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. However, independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000]. Similarly, quantitative analysis of the frequency of positive results (ie. full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@PositiveResultsIncrease]. The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].

## Methods and materials

To objectively quantify the relative hardness or softness of archaeology, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using parameters of scholarly publications. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, number of cited references, age of references, and diversity of references for a large sample of peer-reviewed journal articles. These parameters are useful because of how they signify consensus in a research community. A larger number of authors on a paper reflects collaboration of people working together on a common goal. Collaborators have specialized roles, each of whom has the ability to study a part of the problem with high accuracy and detail, with harder fields having larger groups of collaborators[@zuckerman1972age]. Reflecting this collaboration group size, harder disciplines tend to have higher average numbers of authors on papers. Article length has an inverse correlation with field hardness. In low-consensus, or softer, fields, papers must be longer to present justification, nuance and contextualization of results. I used page numbers to measure article length. Similarly, the number of cited references will typically be higher in low-consensus disciplines where relatively fewer claims are taken for granted and thus many claims need citations for support, relative to harder fields [@skilton2006comparative]. The age of works cited has long been used as a measure of a field's hardness [@moed1998new; @bornerAtlasScience2010], based on the assumption that harder fields assimilate new results more rapidly that softer fields [@price1970citation]. I calculated the Price index for each article, which is the proportion of all cited works that were published in the five years preceding the paper.

```{r}
library(tidyverse)

items_df <- read_rds("data/wos-data-df.rds")

n_articles <- nrow(items_df)
year_max <- max(items_df$year)
year_min <- min(items_df$year)
```

I sampled `r n_articles` papers published during `r year_min`-`r year_max`. This sample was collected from Clarivate's Web of Science database by first selecting the Web of Science category 'Archaeology' and the Document type 'article' (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate's Journal Citation Indicator. Finally I excluded journals with less than 100 articles in the database.

The entire R code [@rcoreteamLanguageEnvironmentStatistical2024] used for all the analysis and visualizations contained in this paper is included in the Supplementary Online Materials at https://doi.org/xxx/xxx/xxx to enable re-use of materials and improve reproducibility and transparency [@marwickComputationalReproducibilityArchaeological2017]. All the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this compendium [@doi:10.1080/00031305.2017.1375986]. The R code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

## Results

### How does archaeology compare to other fields?

```{r}
library(ggrepel)

boxlplot_n_authors <- 
items_df %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             log(authors_n))) +
  geom_boxplot(colour = "red", 
               size = 1)  +
  # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.4,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.4,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.8,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 0.0,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 0.0,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of authors (ln)") +
  xlab("Collaborator group size")

boxlplot_n_pages <- 
items_df %>% 
  ggplot(aes(1,
             log(pages_n))) +
  geom_boxplot(colour = "red", 
               size = 1)  +
   # median for physics, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 1.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 1.8,
                                label = "p"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  # median for soc sci, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.5,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.5,
                                label = "s"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
    # median for humanities, from Fanelli & Glänzel Fig 2
  geom_hline(yintercept = 2.8,
             colour = "grey70") +
  geom_text_repel(data = tibble(x = 1, 
                                y = 2.8,
                                label = "h"),
                  aes(x, y, label = label),
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal() + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of pages (ln)") +
  xlab("Article length")

library(cowplot)

plot_grid(boxlplot_n_authors, 
          boxlplot_n_pages)
```

### How has the hardness of archaeology varied over time?

### How do archaeology journals vary in hardness?

```{=html}
<!-- n an ideal science, scholars share a common background of established theories, facts and methods. This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research. Harder sciences are hypothesised to come closer to this ideal. Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)

@@

My motivation is the view that It is customary in significant round-number anniversary -->
```
